Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2014-11-14 response-to-revision-comments.tex

The authors check for the reliability of their model to recover their parameters by using a 
simulation study. The simulation study covers a large range of different population 
dynamics as well as additional cases where the simulated data deviates from the assumption 
of the model used for the analysis. In addition, to the simulation study, the models are 
applied to real data.

\vspace{0.5cm}
\textit{We thank the reviewer for the positive comments!}
\vspace{0.5cm}

In a general way, effort should be put on tightening the text and better focus
on the main goal of the paper, that is about including models of population
growth in a framework that account for imperfect detection of individuals. The
choice of the model of population growth might be subjective and/or motivated by
the knowledge of the life history of the species under study. Even thought, some
studies, proposed some procedures based on model selection to discriminate among
the different models of population growth. Hence, the Ms would benefit of a
better description of the underlying ecological assumptions of the different
models of population growth. For example, the exponential model describes
population growth in an unlimited environment, where $r$ is the maximum per capita
of increase, etc… . In Ricker model, the density-dependence becomes stronger at
higher densities, due to the exponential function relating to carrying capacity.
The Gompertz model is similar to the Ricker model except that there is a
constant linear decrease in the instantaneous growth rate ($r$) as the natural
logarithm of population increase, etc…..

\vspace{0.5cm}
\textit{We have described the classical population growth models in more 
detail (lines 172--174, 180--182, and 184--186).}
\vspace{0.5cm}

@@

Even if the Dail and Madsen model also provides a structure to account for time dependence
as well as for ``process errors'' through the birth and death process and for imperfect
detection, the Dail and Madsen model and state-space models are defined in two different
``statistical worlds''. Hence, the analytical results of the Dail and Madsen model and a state-space
model would be different. That is why, in my point view, it is unclear and misleading to
relate the Dail and Madsen model to a state-space model.

\vspace{0.5cm}
\textit{Our definition of a state-space model is in agreement with the brand new
book on the subject \citep{newman_etal:2014}. They cover many state-space 
models that have
a first-order Markov state process and a non-Gaussian
observation process. For this reason, we think that our title connects
our work to the bigger picture, which is state-space
models of count data.} 
\vspace{0.5cm}

Lines 118 - 174: the description of the model could be shortened (or moved in part in the
appendices) because it has been described now in further publications.

\vspace{0.5cm}
\textit{Good suggestion!  We have shortened the DM model description section in a few places
where it would not interfere with the descriptions of our extensions.}
\vspace{0.5cm}

@@

Line 135: typography error in \textit{``The third process describes the relationships between''} should
be \textit{``The third process describes the relationships between.''}

\vspace{0.5cm}
\textit{We have fixed the typographic error (surplus ``the'') here.}
\vspace{0.5cm}

Line 190: More explanations on the Gompertz model should be added here, especially on
the link between the Ricker model and the Gompertz model. The original reference of the
Gompertz logistic growth equation should be added here.

\vspace{0.5cm}
\textit{We have described the Gompertz model in more 
detail and added the original reference (lines 182--186).}

@@

\vspace{0.5cm}

Line 199: I am unclear about the meaning of \textit{``internal and external (immigration)
contributions to population growth.''} Then, the authors proposed to account for immigration
but they do not mention anything about emigration. In the Dail and Madsen model, the
survival rate is a reduced demographic parameter that accounts for both death and
emigration. Some words, in the discussion, about how emigration is (or not) handled in this
novel extension would be welcome.

\vspace{0.5cm}
\textit{We have edited the text for clarity (lines 190--194).  We have added discussion about emigration
(lines 464--465 and 473--474).}
\vspace{0.5cm}

@@

Line 225: I am unclear about \textit{``semelparous organisms with one generation per sampling
interval (year).''} Semelparous refers to organisms that died after reproduction, (i.e. they can
have solely one single reproduction event in their life because they die after this
reproduction event). So, semelparous organisms cannot have ``one generation per sampling
interval'' but only one generation.

Lines 225 - 229 – This paragraph needs to be clarified – semelparous organisms cannot have
more than one generation per year because they can have only one single reproduction
event in their life (see above).

\vspace{0.5cm}
\textit{We were referring to generations of the population or species, not the individual.  We have edited 
this paragraph for clarity (lines 216--227).}
\vspace{0.5cm}

@@

Lines 281 - 296: The paragraph about the simulation of the data needs to be clarified to
understand better how the data are simulated. Some references to the different equations
of the model should help to clarify this section. For the simulation study, the time series
which are simulated are rather long (40 years) and also the length of the time series of the
application are about 40 years. If the time series were shorter, how the estimation of the
parameters would be affected.

\vspace{0.5cm}
\textit{We have added clarification and equation references to this paragraph (lines 277--292). 
We expect that
the models would not perform as well with far fewer years of data, and have commented on
this in the discussion (lines 485--486).}
\vspace{0.5cm}

@@

Fig. 1: for initial abundance on x-axis, replace little lambda with big lambda
to match text.

\vspace{0.5cm}
\textit{This figure is no longer in the text.}
\vspace{0.5cm}

Fig. 2 legend, in parentheses could replace ``mean and'' with +/-, given that it
is already stated that the panel shows the mean.

\vspace{0.5cm}
\textit{We have made the suggested change (now Fig. 1 legend).}
\vspace{0.5cm}

@@

Line 358: replace 'inaccurate' with 'imprecise'.

\vspace{0.5cm}
\textit{These terms have multiple meanings and are defined differently in
different places.  We like the definitions of \citet{williams_etal:2002}:
``The tendency of replicated estimates to be dispersed is an expression
of estimator precision,'' and
``Accuracy combines both bias and precision in an assessment of estimator
performance.''  On the same pages (44--45) those authors describe mean squared error 
(MSE) as a measure of estimator accuracy.  Therefore root mean squared error (RMSE) and
relative RMSE are best described as measures of accuracy as well.  
In a few other cases (lines 348 and 365) we used the
word ``imprecise'' when the RMSE was high, but in those cases the estimates were highly dispersed
(Figs. A1 and A4).  In this case, since the estimates for $r$ are not very dispersed (Fig. A2), 
we feel it is more correct to describe
the estimates as ``inaccurate.''}
\vspace{0.5cm}

@@

Lines 424-429: Could difficulty in estimating vital rates derives from having
survival and recruitment parameters taking similar values? For example many
small songbirds have apparent survival rates of $\sim .5$; for a stable population,
recruitment would be similar, which in lieu of direct information on vital
rates, might make the 2 parameters difficult to separate. In the American Robin
example in Dail and Madsen (2011), survival and recruitment were estimated at
reasonable values of .7 and .5 (although precision was still fairly low). Could
low detection probability have something to do with this problem? From the p
estimates in Fig. B1, detection probabilities were $< .2$ (which seems low to me
given that ovenbirds are loud and sing a lot, although I suppose it has to do
with large sample radius). In the simulations in the Dail-Madsen paper, lowest
coverage for gamma and omega was found at lowest p values (p = .25). I realize
I'm stretching a bit here, but it was a little
unsatisfying to me that the vital rates parameters could not be estimated, and
the reason for why this was so wasn't too clear. The authors suggest individual
heterogeneity might be an issue and suggest incorporating prior information or
data via integrated population models to improve vital rate estimation. However,
I wonder if some additional conjecture might be added here?

\vspace{0.5cm}
\textit{We have added some conjectures in this direction (lines 425--427).}
\vspace{0.5cm}

### 2014-11-29 improved_state_space.tex
Final revisions before resubmit
Making normal distribution equations consistent

A simple state-space model can be described as follows.
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and let $X_t$ be
the observed data, which differs from $N_t$ due to observation error,
a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
modeled using a population growth model, $\mu(N_{t-1})$
and random process variation denoted $\eta_t$.
The growth model may be density-dependent, as in the case of the 
Ricker-logistic model (henceforth Ricker model), or it might be 
density-independent, such as when growth
is exponential. 
A generic model can be written: 
\begin{linenomath*}
\begin{gather}
  \label{eq:ss1}
  \begin{align}
    N_1 &= x_0 \nonumber \\ 
N_t &= \mu(N_{t-1}) + \eta_{t-1} \quad \text{for} \; 
t=2,\hdots,T  \\
X_t &= N_t + \zeta_t \qquad \qquad \;\, \text{for} \;
t=1,\hdots,T. \nonumber 
  \end{align}
\end{gather}
\end{linenomath*}
where $x_0$ is the distribution for initial abundance.  In classical
state-space models, the two sets of random effects
are assumed to be i.i.d. Gaussian variables: 
$\eta_t \sim \mathrm{N}(0, \sigma)$ and
$\zeta_t \sim \mathrm{N}(0, \tau)$. 
The process variation associated with $N_1$ is often ignored ($x_0 = X_1$), or
the population is assumed to be at stochastic equilibrium such that the $x_0$ can
be regarded as an outcome of the equilibrium distribution \citep{devalpine_hastings:2002}.

Even though state-space models such as that shown in Eq.~\ref{eq:ss1}
overcome many of the limitations associated with the application of
classical models of population dynamics, 
several problems exist. 
The assumption that random effects are Gaussian distributed makes it
easier to estimate parameters because methods such as the Kalman
filter can be applied 
\citep{dennis_etal:2006}; however, 
these models effectively assume that abundance can be take on any
real value, and hence they can predict negative values of abundance. 
Transforming the count data and the underlying abundance variables 
do little to resolve these issues \citep{ohara_kotze:2010}.

@@

The DM model is an extension of the 
\citet{royle:2004biom} $N$-mixture model, which allows for inference about spatial
variation in abundance when individuals cannot be detected with
certainty. To estimate both abundance and detection parameters, 
the $N$-mixture model uses replicate % original
observations at each site. The observations are made during sufficiently
short time intervals such that the population can be %safely
assumed to be closed with respect to births, deaths, and movement. The DM
model relaxes this closure assumption and includes explicit parameters
describing population change over time.

@@

Like traditional state-space models, the DM model includes 
three conditionally related processes corresponding to: 
(1) initial abundance, i.e. the
abundance at site $i$ during the first primary period,
denoted $N_{i,1}$, (2) abundance at time $t$ (for $t>1$) which depends upon
abundance at $t-1$, and (3) the
detection process. % \citep{dail_madsen:2011}.
The first two processes describe the %the state process---
variation in abundance in space and time. The third process 
describes the relationship between
abundance and the observed count data.

@@

Conventional state-space models assume that the
distribution for the initial time period 
is either the equilibrium
distribution, or has zero variance.
In contrast, \citet{dail_madsen:2011} proposed modeling $N_{i,1}$
%as either a Poisson or negative binomial random variable:
%\begin{linenomath*}
%\begin{gather}
%N_{i,1} \sim \mathrm{Pois}(\Lambda) \nonumber \\
%\text{or} \nonumber \\
%N_{i,1} \sim \mathrm{NB}(\Lambda, \alpha)
%\label{eq:N1}
%\end{gather}
%\end{linenomath*}
%where $\Lambda_i$ is the expected abundance at site $i$ during
%year 1.
as either a Poisson ($N_{i,1} \sim \mathrm{Pois}(\Lambda_i)$) or 
negative binomial random variable ($N_{i,1} \sim \mathrm{NB}(\Lambda_i, \alpha)$),
where $\Lambda_i$ is the expected abundance at site $i$ during
year 1.
The Poisson distribution assumes that the mean of $N_{i,1}$ is
equal to its variance, and the negative binomial distribution allows the
variance to be greater than the mean with the amount of
overdispersion determined by the parameter $\alpha$.

Regardless of the specified distribution, the model for initial
abundance has two distinguishing features. First, it provides a
mechanism for characterizing spatial variation in abundance. For
instance, one could model the effect of an environmental
covariate ($x_i$) on abundance using a log-linear
model such as $\log(\Lambda_i) = \beta^{\Lambda}_0 +
\beta^{\Lambda}_1
x_{i}$. Second, the spatial
replicates resolve the
problem of parameter non-identifiability that are common to
standard state-space models because
process variation and observation error can be estimated from
spatially-replicated count data \citep{royle:2004biom}. %Hence, the first component of
%the model addresses both the issues of spatial inference and
%parameter identifiability discussed previously.

@@

The DM model assumes that abundance at time $t>1$ is a function of
abundance at time $t-1$, i.e. abundance at each site evolves as a
first order Markov process. 
\citet{dail_madsen:2011} considered several models to describe the temporal dynamics;
however, in each case they modeled $N_t$ as the sum of two random variables:
$S_{i,t}$, the number of individuals surviving from $t-1$ and not
emigrating; and $G_{i,t}$ the number of new individuals entering
the population. Their most general model was
\begin{linenomath*}
\begin{equation}
\left.\begin{aligned}
S_{i,t}|N_{i,t-1} &\sim \mathrm{Bin}(N_{i,t-1}, \omega) \\
G_{i,t}|N_{i,t-1} &\sim \mathrm{Pois}(\gamma(N_{i,t-1})) \\
N_{i,t} &= S_{i,t}+G_{i,t}
\end{aligned}\right\} \quad \text{for} \; t=2,\hdots,T
\label{eq:Nt}
\end{equation}
\end{linenomath*}
where $\omega$ is the apparent survival probability and $\gamma$
is the recruitment rate, which can depend on $N_{i,t-1}$.
\citet{dail_madsen:2011} proposed three
models for $\gamma$: the constant model,
$G_{i,t} \sim \mathrm{Pois}(\gamma)$ where recruitment does not
depend on $N_{i,t-1}$; %and which simulates a ``propagule rain'' of new individuals; 
the ``autoregressive'' model, $G_{i,t} \sim
\mathrm{Pois}(\gamma N_{i,t-1})$
of geometric or density independent growth; and the
no-trend model, $\gamma = (1-\omega)\Lambda$, which assumes the
population is at equilibrium. Covariates of
$\omega$ and $\gamma$ can be accommodated, e.g., 
using logit- and log-linear models respectively.

@@

\section*{Model Extensions}
\label{sec:ext}
\subsection*{Population growth models}
Partitioning population growth into survival and recruitment
is useful  
in terms of providing a mechanistic description of population
dynamics, but this is not always possible using simple count data,
especially when the sites are not closed with respect to movement.
%immigration and emigration. 
When the mechanistic model is unrealistic, we suggest replacing it
with classical population growth models that have a long history in ecology.
The exponential growth model, which describes population growth in an 
unlimited environment, is the simplest: 
$N_{i,t} = N_{i,t-1}e^r$ where $r$ is the intrinsic or maximum per capita
rate of increase. 
To allow for demographic stochasticity, we can regard $N_{i,t}$ as a Poisson 
random variable, simplifying Eq.~\ref{eq:Nt} to:
\begin{linenomath*}
\begin{equation}
  N_{i,t} \sim \mathrm{Pois}(\exp(r)N_{i,t-1}).
\label{eq:exp}
\end{equation}
\end{linenomath*}

Density-dependent versions of the model are also possible.  For
example:
\begin{linenomath*}
\begin{equation}
  N_{i,t} \sim \mathrm{Pois}(N_{i,t-1}\exp(r(1-N_{i,t-1}/K)))
\label{eq:rick}
\end{equation}
\end{linenomath*}
where $K$ is the stable equilibrium of the population and $r$ is
the instantaneous population growth rate at low population
densities, and both parameters are constrained to be positive. This is
a stochastic version of the 
Ricker model \citep{ricker:1954}. In the Ricker model, 
density dependence becomes stronger as abundance increases well above $K$,
leading to overcompensation    
%due to the exponential function relating to carrying capacity 
\citep[abundance goes from above $K$ to below it;][]{sabo_etal:2004}.
Another option is a 
modified Gompertz-logistic
density-
dependent model \citep[henceforth Gompertz model;][]{gompertz:1825,
hart_gotelli:2011}:
\begin{linenomath*}
\begin{equation}
N_{i,t} \sim \mathrm{Pois}(N_{i,t-1}\exp(r(1-\log(N_{i,t-1}+1)/\log(K+1))))
\label{eq:gomp}
\end{equation}
\end{linenomath*}
Here the interpretations of $r$ and $K$ are similar to those in the
Ricker model, but dynamics can only be overcompensatory at high values of $r$
(approximately $r > log(K)$).
%population growth declines linearly with an increase in the natural logarithm of abundance,
%rather than linearly with an increase in abundance. 
Because a single Poisson distribution controls
the distribution of $N_{i,t}$ in each of these models, the discrete 
convolution used by \citet{dail_madsen:2011} to construct the
likelihood is not required, speeding up processing time.

@@

Another way to model environmental stochasticity is as lognormal variation in 
the population growth rate \citep{bjornstad:2001,bonsall_hastings:2004}.  
This could be applied to both density-independent and density-dependent models.
For example, for the Ricker + immigration model:
\begin{linenomath*}
\begin{equation}
N_{i,t} \sim
\mathrm{Pois}(N_{i,t-1}\exp(\nu_{i,t} + r(1-N_{i,t-1}/K)) + \iota)
\label{eq:nuRand}
\end{equation}
\end{linenomath*}
where $\nu$ is a variable with a normal distribution, mean value of 0, and standard deviation $\sigma_\nu$.  
We suggest three variations for spatial variation in environmental stochasticity.  
The first has independent environmental stochasticity 
between sites, and might be appropriate where sites are widely dispersed.  In the second, 
all sites have the same environmental conditions in a given year (regional stochasticity, $\nu_{i,t} = \nu_{t}$),
which might be appropriate where the geographic or environmental range is small \citep{hanski:1998}.  
Third, where sites can be close together but the range of sites is broad, a multivariate normal distribution 
for $\nu_{i,t}$ with distance-dependent correlation coefficients 
between sites in a year might be appropriate.

@@

Current BBS trend estimators deal with this problem by
treating observer identity as a random 
effect \citep{link_sauer:2002,sauer_link:2011}.
To include random observer effects in DM models, 
Eq.~\ref{eq:p1} can be modified to:
\begin{linenomath*}
\begin{gather}
X_{i,j,t} \sim \mathrm{Bin}(N_{i,t}, p_j) \nonumber \\
\mathrm{logit}(p_j) \sim \mathrm{N}(\mu_p, \sigma_p)
\label{eq:pobs}
\end{gather}
\end{linenomath*}
where $X_{i,j,t}$ is the number of individuals counted at site $i$ by
observer $j$ in year $t$, $p_j$ is observer-specific detection probability,
$\mu_p$ is the mean detection probability (on the logit scale), and $\sigma_p$ is
the logit-scale standard deviation of the random observer effect.

@@

Although we view this as an important connection to make between the two
different classes of state-space models, we believe that
such population growth models are more phenomenological
than mechanistic, and this runs counter to the motivation for
the DM model. However, estimating
demographic parameters from count data is an ambitious goal, and
the required assumptions will not be valid in many cases,
especially when immigration and emigration occur. This was supported by the
unrealistic estimates of vital rates reported here and by the
lack of support for the mechanistic models relative to the population
growth models.  A separate analysis of BBS data for eleven other forest passerines
found similar poor support and unrealistic estimates for the mechanistic 
models (J. Hostetler, unpublished analyses), suggesting that some process, possibly
individual heterogeneity \citep{vindenes_etal:2008}, is increasing variances
above the levels expected from demographic stochasticity alone.  It
is also possible that similar values for recruitment and survival rates in songbirds and low 
detection probabilities for the BBS have contributed %from the large sample radii 
to the DM model issues.

%One strategy for adhering to the original formulation of the
%model that included survival and recruitment parameters would be to
%use either prior information or additional data \citep{zipkin_etal:2014}. For example, in a 
%Bayesian analysis existing estimates of demographic parameters
%could be used as informative priors, which might make it possible to
%seperate the effects of movement versus births and deaths.
%Alternatively, ecologists could combine count data with demographic data,
%often referred to as an integrated population model
%\citep[IPM;][]{besbeas_etal:2002,schaub_etal:2007,chandler_clark:2014}.
%We believe that a DM modeling framework could work well in this context,
%and that the DM model could be viewed as a specific case of the
%IPM proposed by \citet{buckland_etal:2004}.
%However, we caution that when count data are affected by
%movement and vital rates, it would be unwise to ignore the movement
%processes.
%
%Integrated population models may be viewed as the gold standard in
%state-space modeling, but ecologists do not always have direct
%information about vital rates, especially at large spatial scales.
%Count data are much more common, relatively inexpensive to obtain, and
%are produced by many of the largest monitoring programs in the
%world, such as the BBS. 
%In the absence of direct information about demographic parameters, and
%when the original assumptions of the model do not hold, our extensions to the
%DM model can still be used for important purposes such as predicting future
%population size and changes in species distributions.

@@

Our emphasis was on increasing the practical utility of this
class of models, and so we avoided several conceptually interesting
extensions that we believe would be computationally prohibitive in many
cases. Nonetheless, we will discuss one --- spatially-explicit models of
immigration. 
In our models, immigration is currently modeled as independent of
population sizes in other sites, and emigration is included only
as an implicit piece of apparent survival or population growth. 
We suggest alternatives that put movement in a metapopulation
context: immigration conditional on total or mean % \citep{hanski:1998}
abundance across sites the previous time step; immigration
conditional on abundance in nearby sites only; and immigration
conditional on abundance at sites as a function of their distance
from the receiving site \citep{hastings:1991,hanski:1998}. These models would
require that all sites in the metapopulation are sampled, or that abundance at 
unsampled locations can be estimated \citep{lele_etal:1998}. 
Initial tests suggest that it would be
possible to fit some or all of these models, at least in the
Bayesian framework. It may also be possible to model an explicit emigration
process or rate and tie immigration at nearby sites to this rate.

>>>>

A simple state-space model can be described as follows.
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and let $X_t$ be the observed data. 
Temporal variation in $N_t$ is
modeled using a probability distribution whose expected value is
determined by a population growth model, $\mu(N_{t-1}, {\bm \theta})$, such as 
a density-dependent Ricker-logistic model (henceforth Ricker model) or  
a density-independent exponential model, with parameter vector $\bm
\theta$. Such a model can be written:  
\begin{linenomath*}
\begin{gather}
  \label{eq:ss1}
  \begin{align}
  N_1 &\sim \mathrm{Normal}(\mu_1, \sigma^{2}_0) \nonumber \\
  N_t &\sim \mathrm{Normal}(\mu(N_{t-1}, {\bm \theta}), \sigma^{2}) \qquad
  \; \text{for} \;  t=2,\hdots,T  \\
  X_t &\sim \mathrm{Normal}(N_t, \tau^{2}) \qquad \qquad \quad \;\;\, \text{for} \;  t=1,\hdots,T  \nonumber
  \end{align}
\end{gather}
\end{linenomath*}
where $\sigma^{2}_0$ and $\sigma^{2}$ are variance parameters of the
underlying abundance process, and $\tau^{2}$ describes variation in 
observation error.

Even though state-space models such as that shown in Eq.~\ref{eq:ss1}
overcome many of the limitations associated with the application of
classical models of population dynamics, several problems exist. 
The Gaussian assumptions 
make it easier to estimate parameters because methods such as the Kalman
filter can be applied \citep{dennis_etal:2006}; however, 
these models allow for implausible preditions such as 
negative values of abundance. 
Transforming the count data and the underlying abundance variables 
do little to resolve these issues \citep{ohara_kotze:2010}.

@@

The DM model is an extension of the 
\citet{royle:2004biom} $N$-mixture model, which allows for inference about spatial
variation in abundance when individuals cannot be detected with
certainty. To estimate both abundance and detection parameters, 
the $N$-mixture model uses replicate 
observations at each site. The observations are made during sufficiently
short time intervals such that the population can be 
assumed to be closed with respect to births, deaths, and movement. The DM
model relaxes this closure assumption and includes explicit parameters
describing population change over time.

@@

Like traditional state-space models, the DM model includes 
three conditionally related processes corresponding to: 
(1) initial abundance, i.e. the
abundance at site $i$ during the first primary period,
denoted $N_{i,1}$, (2) abundance at time $t$ (for $t>1$) which depends upon
abundance at $t-1$, and (3) the
detection process. 
The first two processes describe the 
variation in abundance in space and time. The third process 
describes the relationship between
abundance and the observed count data.

@@

Conventional state-space models assume that the
distribution for the initial time period 
is either the equilibrium
distribution, or has zero variance.
In contrast, \citet{dail_madsen:2011} proposed modeling $N_{i,1}$
as either a Poisson ($N_{i,1} \sim \mathrm{Pois}(\Lambda_i)$) or 
negative binomial random variable ($N_{i,1} \sim \mathrm{NB}(\Lambda_i, \alpha)$),
where $\Lambda_i$ is the expected abundance at site $i$ during
year 1.
The Poisson distribution assumes that the mean of $N_{i,1}$ is
equal to its variance, and the negative binomial distribution allows the
variance to be greater than the mean with the amount of
overdispersion determined by the parameter $\alpha$.

Regardless of the specified distribution, the model for initial
abundance has two distinguishing features. First, it provides a
mechanism for characterizing spatial variation in abundance. For
instance, one could model the effect of an environmental
covariate ($x_i$) on abundance using a log-linear
model such as $\log(\Lambda_i) = \beta^{\Lambda}_0 +
\beta^{\Lambda}_1
x_{i}$. Second, the spatial
replicates resolve the
problem of parameter non-identifiability that are common to
standard state-space models because
process variation and observation error can be estimated from
spatially-replicated count data \citep{royle:2004biom}.

@@

The DM model assumes that abundance at time $t>1$ is a function of
abundance at time $t-1$, i.e. abundance at each site evolves as a
first order Markov process. 
\citet{dail_madsen:2011} considered several models to describe the temporal dynamics;
however, in each case they modeled $N_t$ as the sum of two random variables:
$S_{i,t}$, the number of individuals surviving from $t-1$ and not
emigrating; and $G_{i,t}$ the number of new individuals entering
the population. Their most general model was
\begin{linenomath*}
\begin{equation}
\left.\begin{aligned}
S_{i,t}|N_{i,t-1} &\sim \mathrm{Bin}(N_{i,t-1}, \omega) \\
G_{i,t}|N_{i,t-1} &\sim \mathrm{Pois}(\gamma(N_{i,t-1})) \\
N_{i,t} &= S_{i,t}+G_{i,t}
\end{aligned}\right\} \quad \text{for} \; t=2,\hdots,T
\label{eq:Nt}
\end{equation}
\end{linenomath*}
where $\omega$ is the apparent survival probability and $\gamma$
is the recruitment rate, which can depend on $N_{i,t-1}$.
\citet{dail_madsen:2011} proposed three
models for $\gamma$: the constant model,
$G_{i,t} \sim \mathrm{Pois}(\gamma)$ where recruitment does not
depend on $N_{i,t-1}$; 
the ``autoregressive'' model, $G_{i,t} \sim
\mathrm{Pois}(\gamma N_{i,t-1})$
of geometric or density independent growth; and the
no-trend model, $\gamma = (1-\omega)\Lambda$, which assumes the
population is at equilibrium. Covariates of
$\omega$ and $\gamma$ can be accommodated, e.g., 
using logit- and log-linear models respectively.

@@

\section*{Model Extensions}
\label{sec:ext}
\subsection*{Population growth models}
Partitioning population growth into survival and recruitment
is useful  
in terms of providing a mechanistic description of population
dynamics, but this is not always possible using simple count data,
especially when the sites are not closed with respect to movement.
When the mechanistic model is unrealistic, we suggest replacing it
with classical population growth models that have a long history in ecology.
The exponential growth model, which describes population growth in an 
unlimited environment, is the simplest: 
$N_{i,t} = N_{i,t-1}e^r$ where $r$ is the intrinsic or maximum per capita
rate of increase. 
To allow for demographic stochasticity, we can regard $N_{i,t}$ as a Poisson 
random variable, simplifying Eq.~\ref{eq:Nt} to:
\begin{linenomath*}
\begin{equation}
  N_{i,t} \sim \mathrm{Pois}(\exp(r)N_{i,t-1}).
\label{eq:exp}
\end{equation}
\end{linenomath*}

Density-dependent versions of the model are also possible.  For
example:
\begin{linenomath*}
\begin{equation}
  N_{i,t} \sim \mathrm{Pois}(N_{i,t-1}\exp(r(1-N_{i,t-1}/K)))
\label{eq:rick}
\end{equation}
\end{linenomath*}
where $K$ is the stable equilibrium of the population and $r$ is
the instantaneous population growth rate at low population
densities, and both parameters are constrained to be positive. This is
a stochastic version of the 
Ricker model \citep{ricker:1954}. In the Ricker model, 
density dependence becomes stronger as abundance increases well above $K$,
leading to overcompensation    
\citep[abundance goes from above $K$ to below it;][]{sabo_etal:2004}.
Another option is a 
modified Gompertz-logistic
density-
dependent model \citep[henceforth Gompertz model;][]{gompertz:1825,
hart_gotelli:2011}:
\begin{linenomath*}
\begin{equation}
N_{i,t} \sim \mathrm{Pois}(N_{i,t-1}\exp(r(1-\log(N_{i,t-1}+1)/\log(K+1))))
\label{eq:gomp}
\end{equation}
\end{linenomath*}
Here the interpretations of $r$ and $K$ are similar to those in the
Ricker model, but dynamics can only be overcompensatory at high values of $r$
(approximately $r > log(K)$).
Because a single Poisson distribution controls
the distribution of $N_{i,t}$ in each of these models, the discrete 
convolution used by \citet{dail_madsen:2011} to construct the
likelihood is not required, speeding up processing time.

@@

Another way to model environmental stochasticity is as lognormal variation in 
the population growth rate \citep{bjornstad:2001,bonsall_hastings:2004}.  
This could be applied to both density-independent and density-dependent models.
For example, for the Ricker + immigration model:
\begin{linenomath*}
\begin{equation}
N_{i,t} \sim
\mathrm{Pois}(N_{i,t-1}\exp(\nu_{i,t} + r(1-N_{i,t-1}/K)) + \iota)
\label{eq:nuRand}
\end{equation}
\end{linenomath*}
where $\nu$ is a variable with a normal distribution, mean value of 0, and variance $\sigma^{2}_\nu$.  
We suggest three variations for spatial variation in environmental stochasticity.  
The first has independent environmental stochasticity 
between sites, and might be appropriate where sites are widely dispersed.  In the second, 
all sites have the same environmental conditions in a given year (regional stochasticity, $\nu_{i,t} = \nu_{t}$),
which might be appropriate where the geographic or environmental range is small \citep{hanski:1998}.  
Third, where sites can be close together but the range of sites is broad, a multivariate normal distribution 
for $\nu_{i,t}$ with distance-dependent correlation coefficients 
between sites in a year might be appropriate.

@@

Current BBS trend estimators deal with this problem by
treating observer identity as a random 
effect \citep{link_sauer:2002,sauer_link:2011}.
To include random observer effects in DM models, 
Eq.~\ref{eq:p1} can be modified to:
\begin{linenomath*}
\begin{gather}
X_{i,j,t} \sim \mathrm{Bin}(N_{i,t}, p_j) \nonumber \\
\mathrm{logit}(p_j) \sim \mathrm{Normal}(\mu_p, \sigma^{2}_p)
\label{eq:pobs}
\end{gather}
\end{linenomath*}
where $X_{i,j,t}$ is the number of individuals counted at site $i$ by
observer $j$ in year $t$, $p_j$ is observer-specific detection probability,
$\mu_p$ is the mean detection probability (on the logit scale), and $\sigma^{2}_p$ is
the logit-scale variance of the random observer effect.

@@

Although we view this as an important connection to make between the two
different classes of state-space models, we believe that
such population growth models are more phenomenological
than mechanistic, and this runs counter to the motivation for
the DM model. However, estimating
demographic parameters from count data is an ambitious goal, and
the required assumptions will not be valid in many cases,
especially when immigration and emigration occur. This was supported by the
unrealistic estimates of vital rates reported here and by the
lack of support for the mechanistic models relative to the population
growth models.  A separate analysis of BBS data for eleven other forest passerines
found similar poor support and unrealistic estimates for the mechanistic 
models (J. Hostetler, unpublished analyses), suggesting that some process, possibly
individual heterogeneity \citep{vindenes_etal:2008}, is increasing variances
above the levels expected from demographic stochasticity alone.  It
is also possible that similar values for recruitment and survival rates in songbirds and low 
detection probabilities for the BBS have contributed 
to the DM model issues.

@@

Our emphasis was on increasing the practical utility of this
class of models, and so we avoided several conceptually interesting
extensions that we believe would be computationally prohibitive in many
cases. Nonetheless, we will discuss one --- spatially-explicit models of
immigration. 
In our models, immigration is currently modeled as independent of
population sizes in other sites, and emigration is included only
as an implicit piece of apparent survival or population growth. 
We suggest alternatives that put movement in a metapopulation
context: immigration conditional on total or mean 
abundance across sites the previous time step; immigration
conditional on abundance in nearby sites only; and immigration
conditional on abundance at sites as a function of their distance
from the receiving site \citep{hastings:1991,hanski:1998}. These models would
require that all sites in the metapopulation are sampled, or that abundance at 
unsampled locations can be estimated \citep{lele_etal:1998}. 
Initial tests suggest that it would be
possible to fit some or all of these models, at least in the
Bayesian framework. It may also be possible to model an explicit emigration
process or rate and tie immigration at nearby sites to this rate.
